---
output: html_document
title: "Phenotype descriptions"
---

```{r setup, include=F}
#command args
args = commandArgs(trailingOnly = TRUE)
ANC = "EUR" #args[1]

knitr::opts_chunk$set(echo = FALSE, message = FALSE,warning = FALSE, fig.path = paste0("../output/phenos_summary/", ANC, "_"))
suppressMessages(silent <- lapply(c("tidyverse", "data.table", "knitr", "kableExtra", "paletteer", "ggpubr"),  library, character.only = TRUE))
palettes <- list(NatComms=paletteer_d("ggsci::nrc_npg", n=10),
                 HeatMapYlRd=rev(paletteer::paletteer_c("grDevices::heat.colors", n=100)))
```


```{r basics}
# load basic functions
source("basic_functions.R", echo=F)

# set color palettes
palettes <- list(NatComms=paletteer_d("ggsci::nrc_npg", n=10),
                 HeatMapYlRd=rev(paletteer::paletteer_c("grDevices::heat.colors", n=100)))
Taster_palette_Labs <- c("Nontaster" = palettes$NatComms[1], "Taster"= palettes$NatComms[4], 
                         "Supertaster"= palettes$NatComms[3], "Other" = "grey50")

# set ggplot theme standards
ggplot_theme_standard_continuous <- theme_bw() + theme(
    axis.text.x = element_text(size=12, vjust=0.65, color = "black"),
    axis.text.y = element_text(size=12, color="black"), 
    strip.text=element_text(size=14, face="bold"),
    axis.title = element_text(size=14, color = "black")
)

ggplot_theme_standard_categorical <- theme_bw() + theme(
    axis.text.x = element_text(size=12, color = "black", angle=30, hjust=0.9),
    axis.text.y = element_text(size=12, color="black"), 
    strip.text=element_text(size=14, face="bold"),
    axis.title = element_text(size=14, color = "black")
)
```

```{r read-phenos}
analysis.prep <- readRDS(paste0("../data/processed/ukb_analysis_", ANC, "_rm5sd.rda"))
```


### Exclusion criteria

* No diabetes (Eastwood algorithm + HbA1c <5.7%)
* Common TAS2R38 diplotype
* [ANC] genetic ancestry
* Complete data for glucose, genetic ancestry, covariates
* Plausible total energy intakes (600-4000 kcal/d)


```{r exclusion-criteria}
analysis.prep %>%
  select(id, starts_with("N_")) %>%
  pivot_longer(-id, names_to="criterion") %>%
  group_by(criterion) %>%
  summarise(Yes = sum(value),
            No = sum(!value)) %>%
  setNames(c("Inclusion criterion", "Yes", "No")) %>%
  kable(caption = paste0(ANC, " sample counts for inclusion criteria")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, position = "left") 
analysis <- analysis.prep %>% filter(N_contrl_taste_kcal_compl == 1)
```


\n
\n

### Variable distributions
```{r descr-fns}
plot_continuous <- function(cont_var) {
  analysis %>% select(var=all_of(cont_var)) %>% filter(!is.na(var)) %>%
    ggplot(aes(x=var)) + geom_histogram(bins=30) +
    labs(title=cont_var, x=cont_var, y="frequency") + 
    theme(plot.title = element_text(face="bold")) #+
    #ggplot_theme_standard_continuous
}

plot_categorical <- function(cat_var) {
  analysis %>% 
    select(var=all_of(cat_var)) %>% filter(!is.na(var)) %>%
    ggplot(aes(x=factor(var))) + geom_bar(stat="count") +
    labs(title=cat_var, x=cat_var) + xlab("") +
    theme(axis.text.x = element_text(angle=35, vjust=0.75)) +
    theme(plot.title = element_text(face="bold"))
    #ggplot_theme_standard_categorical
}

```

```{r plot-phenos-lifestyle, fig.height=6, fig.width=12}
ggarrange(
  ggarrange(plot_continuous("age"), plot_categorical("sex"), plot_continuous("bmi"),  
            align="h", nrow=1, ncol=4, widths = c(.5, 0.5, 0.5)),
  ggarrange(plot_categorical("smoke_level.lab"),  plot_categorical("pa_met_excess_lvl"), plot_categorical("alch_freq.lab"), 
            ncol=4, widths=c(0.75, 0.75, 1, 0.5), align = "h"), 
  nrow=2, heights = c(0.75, 1))
```

```{r plot-phenos-diet, fig.width=12, fig.height=6}
ggarrange(plot_continuous("CHO2FIB"), plot_continuous("dietPC1"), plot_continuous("dietPC2"), "",
          plot_continuous("dietPC3"), plot_continuous("dietPC4"), plot_continuous("dietPC5"), 
          ncol=4, nrow=2, widths=c(0.65, 0.65, 0.65, 0.5)
)
```

\n
\n

## Descriptive characteristics by Taster Status

```{r, diplo-freq, fig.height=4, fig.width=8}
# Use dataset with all available diplotypes, among included participants
diplo_gt05=names(which(prop.table(table(
  (analysis.prep %>% filter(incl_t2d==1 & incl_kcal==1 & incl_compl==1))$diplo))>0.0005))

as.data.frame(table((analysis.prep %>% filter(incl_t2d==1 & incl_kcal==1 & incl_compl==1))$diplo)) %>% 
  filter(Var1 %in% diplo_gt05) %>%
  mutate(#order=1:nrow(.), 
         color = ifelse(Var1 == "CAT/CAT", "Nontaster",
                        ifelse(Var1 %in% c("CAT/GGC", "GGC/CAT"), "Taster",
                               ifelse(Var1 == "GGC/GGC", "Supertaster", "Other")))) %>%
  mutate(diplo_order=factor(Var1, levels=Var1[order(Freq, decreasing=T)])) %>%
  ggplot(aes(x=diplo_order, y=Freq/sum(Freq), fill=color)) + 
  geom_bar(stat = "identity") + 
  scale_x_discrete(name = "\n TAS2R38 Diplotypes") +
  ylab("Frequency (%)") + #ylim(0, 0.5) +
  scale_fill_manual(values = Taster_palette_Labs, name="Taster Status") + 
  theme(axis.text.x = element_text(size=10, angle=35, vjust=0.65, color = "black"),
        axis.text.y = element_text(size=10, color = "black"),
        axis.title = element_text(color = "black", face = "bold")) +
  geom_text(aes(label=Freq), vjust=-0.15, size=3) + 
  ggtitle("TAS2R38 Diplotype Frequency Table") 

```

```{r descr}
tab_taste_descr <- analysis %>% filter(complete.cases(taster_status)) %>%
  group_by(taster_status) %>%
  reframe(
    Age=mean_sd(age),
    Female=n_pct(sex, level = "Female"),
    BMI=mean_sd(bmi),
    Smoke_Current=n_pct(smoke_level.lab, level="Current"),
    Smoke_Previous=n_pct(smoke_level.lab, level="Previous"),
    Smoke_Never=n_pct(smoke_level.lab, level="Never"),
    PA_Low=n_pct(pa_met_excess_lvl, level="Low"),
    PA_Moderate=n_pct(pa_met_excess_lvl, level="Moderate"),
    PA_High=n_pct(pa_met_excess_lvl, level="High"),
    Alchohol_Daily=n_pct(alch_freq.lab, level="Daily or almost daily"),
    Alcohol_3to4wk=n_pct(alch_freq.lab, level="3-4 per week"),
    Alcohol_1to2wk=n_pct(alch_freq.lab, level="1-2 per week"), 
    Alcohol_1to3mnth=n_pct(alch_freq.lab, level="1-3 per month"), 
    Alcohol_occsions=n_pct(alch_freq.lab, level="Special occasions only"), 
    Alcohol_never=n_pct(alch_freq.lab, level="Never"),
    Glucose=mean_sd(glu),
    HbA1c=mean_sd(hba1c_max),
    SBP=mean_sd(sbp),
    DBP=mean_sd(dbp),
    Triglyceride=mean_sd(tg),
    LDL=mean_sd(ldl),
    HDL=mean_sd(hdl)) %>% 
  t() 

tab_taste_descr %>% kable(caption = paste0(ANC, " participant characteristics by taster status")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, position = "left")

# Glucose/A1v by fasting time 
tab_glua1c_byFast <- bind_rows(
  analysis %>% filter(complete.cases(taster_status)) %>%
    group_by(fast_cat, taster_status) %>%
    reframe(Glucose=mean_sd(glu)) %>%
    pivot_wider(names_from="taster_status", values_from="Glucose") %>%
    mutate(Biomarker="Glucose", .before = fast_cat),
  analysis %>% filter(complete.cases(taster_status)) %>%
    group_by(fast_cat, taster_status) %>%
    reframe(HbA1c=mean_sd(hba1c_max)) %>%
    pivot_wider(names_from="taster_status", values_from="HbA1c") %>%
    mutate(Biomarker="HbA1c", .before = fast_cat)
)

tab_glua1c_byFast %>% kable(caption = "Table 3. Mean (SD) T2D biomarkers by taster status") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, position = "left") 
```

\n
\n

#### Diet characteristics
```{r diet-patterns, fig.width=12, fig.height=8}
dietPC.dat <- readRDS(paste0("../data/processed/diet/ukb_", ANC, "_dietPCs.rda"))
as.data.frame(dietPC.dat$rotation) %>%
  mutate(diet=rownames(.)) %>%
  select(diet, c(paste0("PC", 1:10))) %>%
  mutate(diet=factor(diet, levels=diet[order(PC1, decreasing=T)])) %>%
  pivot_longer(-diet) %>% 
  mutate(name=factor(name, levels=c(paste0("PC", 1:10)) )) %>%
  mutate(direction = ifelse(value <=0, "neg", "pos")) %>%
  ggplot(aes(x=value, y=diet, fill=direction)) + 
  facet_wrap(~name, ncol=5) + 
  #geom_vline(xintercept = 0, color = "black") +
  geom_col() + ylab("Diet traits derived from UKB FFQ") + xlab("Rotated Factor Loadings") +
  geom_vline(xintercept = c(-0.2, 0.2), color = "black", linetype = "dashed") +
  geom_vline(xintercept = 0, color = "black") +
  scale_fill_manual(values=c(palettes$NatComms[7], palettes$NatComms[5]), 
                    name = "Direction of\nFactor Loading", labels = c("Negative", "Positive")) +
  theme_bw() +
  theme(panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank())
```

```{r diet-descr}
dietVars <- c("TCALS", "CHO_pct", "FIBER", "CHO2FIB", "FAT_pct", "PRO_pct", "ALC", paste0("dietPC", 1:10))
tab_taste_diet <- as.data.frame(analysis %>% filter(complete.cases(taster_status)) %>%
  group_by(taster_status) %>%
  reframe(
    Energy_kcal=mean_sd(TCALS),
    Carbohydrate_pct = mean_sd(CHO_pct),
    Fiber_g = mean_sd(FIBER),
    Carb_to_Fiber=mean_sd(CHO2FIB),
    Fat_pct = mean_sd(FAT_pct),
    Protein_pct = mean_sd(PRO_pct), 
    Alcohol_g=mean_sd(ALC),
    Diet_Pattern_PC1 = mean_sd(dietPC1),
    Diet_Pattern_PC2 = mean_sd(dietPC2),
    Diet_Pattern_PC3 = mean_sd(dietPC3),
    Diet_Pattern_PC4 = mean_sd(dietPC4),
    Diet_Pattern_PC5 = mean_sd(dietPC5),
    Diet_Pattern_PC6 = mean_sd(dietPC6),
    Diet_Pattern_PC7 = mean_sd(dietPC7),
    Diet_Pattern_PC8 = mean_sd(dietPC8),
    Diet_Pattern_PC9 = mean_sd(dietPC9),
    Diet_Pattern_PC10 = mean_sd(dietPC10)) %>% 
  t()) %>% 
  mutate(P_Ftest=c(NA, sapply(dietVars, function(y) {
  anova(lm(formula(paste0(y, "~taster_status+age+sex")), 
           data=analysis %>% filter(taster_status != "Other")))$`P`[1]} ))) %>%
  rename("Nontasters"=V1, "Tasters"=V2, "Supertasters"=V3) %>% filter(rownames(.) != "taster_status")

tab_taste_diet %>%
  kable(caption=paste0(ANC, "participant diet characteristics by taster status")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, position = "left")
```

```{r write-csvs}
#descriptives
as.data.frame(tab_taste_descr) %>% 
  rename("Nontasters"=V1, "Tasters"=V2, "Supertasters"=V3) %>% 
  filter(rownames(.) != "taster_status") %>% 
  write.csv(paste0("../output/phenos_summary/", ANC, "_descrByTaste.csv"))

tab_glua1c_byFast %>% 
  write.csv(paste0("../output/phenos_summary/", ANC, "_gluA1cByTasteXFast.csv"))

tab_taste_diet %>% 
  write.csv(paste0("../output/phenos_summary/", ANC, "_dietByTaste.csv"))


```

##EOF



    
    





