---
title: "results_summary"
output: html_document
date: "2024-06-13"
---

```{r setup, include=FALSE, echo=F, error=F, warning=F, message=F}
knitr::opts_chunk$set(echo = F, error = F, warning = F, out.width = "90%", out.height = "50%")

lapply(c("tidyverse", "data.table", "paletteer", "ggpubr", "gplots",
         "R3port", "xtable", "knitr", "tinytex", "emmeans"),  library, character.only = TRUE)

source("../scripts/pantry.R", echo=F)

```


```{r}
###############################
##   Load data by ancestry   ##
###############################

# system arguments
EUR=c("European"="EUR")

#load data for EUR ancestry
EUR.df <- readRDS(paste0("../data/processed/ukb_analysis_", EUR, ".rda")) %>%
  mutate(taster_status = factor(taster_status, levels=c("Nontaster", "Taster","Supertaster", "Other"))) 

#load dietPCs
dietPCs.dat <- readRDS(paste0("../data/processed/diet/ukb_EUR_dietPCs.rda"))
dietPC_emm <- read.csv("../data/processed/analysis/EUR_emm_dietPC_diplos_sexage.csv")
dietPC_pstat <- read.csv("../data/processed/analysis/EUR_pstat_dietPC_diplos_sexage.csv")
```

\n

## *Inclusion/Exclusion criteria*
* No diabetes data (Eastwood algorithm + HbA1c <5.7%)
* Complete data for glucose, genetic ancestry, and behavioral/SES covariates
 

Planned sensitivity analyses
* Restrict to individuals with plausible total energy intakes : <20MJ for men; <18 MJ for women (range based on values considered plausible by UKB in DataShowcase with additional exclusions to those with >600 kcal, as a lower limit)

```{r incl-anc}

# Sample inclusions
EUR.df %>% 
  mutate(N_geno = ifelse(is.na(haplo_0) == F & is.na(haplo_1) == F, "1", "0")) %>%
  mutate(N_fasting = ifelse(fasting_hrs<=24, "1", "0")) %>%
  mutate(N_contrl_compl_fast=ifelse(N_contrl_compl==1 & N_fasting == 1,"1","0")) %>%
  #filter(N_geno == 1) %>%
  select(starts_with("N_")) %>% 
  reframe(N_European=n_pct(N_Ancestry, level="1"),
          N_geno=n_pct(N_geno, level="1"),
          N_contrl=n_pct(N_contrl, level="1"),
          N_fasting_lt24hr=n_pct(N_fasting, level="1"),
          N_contrl_compl=n_pct(N_contrl_compl, level="1"),
          N_contrl_compl_fast=n_pct(N_contrl_compl_fast, level="1"),
          N_contrl_compl_kcal=n_pct(N_contrl_compl_kcal, level="1"),
          N_contrl_compl_kcal_taste=n_pct(N_contrl_compl_kcal_taste, level="1")) %>% 
  t() %>%
  kable()

analysis <- EUR.df %>% 
  filter(N_contrl_compl == 1) %>%
  filter(is.na(haplo_0) == F & is.na(haplo_1) == F) %>%
  filter(fasting_hrs <= 24) 

dim(analysis) #N=241,955
```
 
  
## Distribution of TAS2R38 Haplotypes & Diplotypes 
We examined the distributions of common (frequency >0.0005) TAS2R38 genotypess 

```{r}

# ================================
## Summary of haplotypes
# ================================

## Add haplo_0_aa and haplo_1_aa to analysis.l dataframe list
analysis <- analysis %>% left_join(geno_to_aa.fun(analysis), by = "id")

## Table
tab_haplos_AA <- analysis %>%
  select(id, haplo_0_aa, haplo_1_aa) %>%
  pivot_longer(-id, values_to="haplo") %>%
  group_by(haplo) %>%
  reframe(n=n(), pct=(n()/nrow(.))*100) %>%
  arrange(n) %>%
  mutate(haplo_order = factor(haplo, levels=haplo[order(n, decreasing = T)])) %>%
  #Filter to "common" haplotypes (>1% of the cohort; >0.005, frequency)
  filter(pct>0.05)

## Plot
plot_haplos_AA <- tab_haplos_AA %>% 
  ggplot(aes(x=haplo_order, y=pct, fill=haplo_order)) + 
  geom_bar(stat = "identity") + 
  scale_x_discrete(name = "TAS2R38 Haplotypes") +
  scale_y_continuous(limits=c(0,62), breaks=seq(0,60,10)) +
  ylab("Frequency (%)") + 
  scale_fill_manual(values = palettes$nature_haplos_yrb, name="TAS2R38 Haplotypes (n)") + 
  ggtheme +
    theme(axis.text.x = element_text(size=8, angle=25, hjust=1, color = "black"), 
          axis.text.y = element_text(size=8, color = "black"),
          axis.title = element_text(color = "black", face = "bold", size=10),
          legend.position = "bottom") +
    geom_text(aes(label=n), vjust=-0.35, size=3.0)

#plot_haplos_AA %>% ggsave(filename="../output/EUR_plot_haplo_005_AA.pdf", height = 4, width=4.5)



# ================================
## Summary of Diplotypes
# ================================

## Create variable in analysis for **common diplotypes**
diplos_gt005_AA <- names(which(prop.table(table(analysis$diplos_common_AA))>0.005))

analysis <- analysis %>%
  mutate(diplos_AA = factor(ifelse(diplos_common_AA %in% diplos_gt005_AA, diplos_common_AA, NA),
                            levels=c(diplos_gt005_AA)))

## Table
tab_diplos_AA <- analysis %>% 
  filter(complete.cases(diplos_AA)) %>% #diplos_common_AA %in% diplos_common_AA) %>% 
  reframe(n=table(diplos_AA), pct=prop.table(table(diplos_AA)), diplos_AA=names(table(diplos_AA))) %>%
  arrange(-n) %>% 
  mutate(color = factor(ifelse(diplos_AA == "AVI/AVI", "Nontaster",
                                 ifelse(diplos_AA == "AVI/PAV", "Taster",
                                        ifelse(diplos_AA == "PAV/PAV", "Supertaster", "Other"))),
                          levels=c("Nontaster","Taster","Supertaster","Other"))) %>%
  mutate(diplos_AA_order=factor(diplos_AA, levels=diplos_AA[order(n, decreasing=T)])) %>%
  #filter to "common" diplotypes (>0.005, frequency)
  filter(pct>0.005)


## Plot 
plot_diplo_AA <- tab_diplos_AA %>%
  ggplot(aes(x=diplos_AA_order, y=pct*100, fill=diplos_AA_order)) + 
  geom_bar(stat = "identity") + 
  scale_x_discrete(name = "TAS2R38 Diplotypes") +
  scale_y_continuous(limits=c(0,52), breaks=seq(0,50,10)) +
  ylab("Frequency (%)") + 
  scale_fill_manual(values = palettes$nature_diplosAA_canonical_yrb, name="TAS2R38 Diplotypes") + 
  ggtheme +
    theme(axis.text.x = element_text(size=8, angle=25, hjust=1, color = "black"), 
          axis.text.y = element_text(size=8, color = "black"),
          axis.title = element_text(color = "black", face = "bold", size=10),
          legend.position = "bottom") +
  geom_text(aes(label=n), vjust=-0.35, size=3.0)

#plot_diplo_AA %>% ggsave(filename="../output/EUR_plot_diplo_005_AA.pdf", height = 4, width=6)


# =================================================
# Panel plot of TAS2R38 haplotypes & diplotypes 
# =================================================
plot_haplo_diplo_AA <- ggarrange(plot_haplos_AA, plot_diplo_AA, ncol=2, widths=c(0.75,1), 
                                 align = c("hv"), labels=c("A", "B"),
          common.legend = T, legend = "none")

plot_haplo_diplo_AA
#plot_haplo_diplo_AA %>% ggsave(filename="../output/EUR_plot_haplo_diplo_005_AA_v3.pdf", height=4, width=9)
 
```



## Descriptive characteristics 
We summarized demographic, behavioral and health characteristics among all identified TAS2R38 diplotypes.

```{r}

## Additional coding of decriptive variables 
analysis <- analysis %>% mutate(
  alch_1to4d.lab = ifelse(alch_freq.lab == "3-4 per week" | alch_freq.lab == "1-2 per week", "1-4 per week", "other"),
  addsalt_3lvl.lab = factor(ifelse(addsalt.lab == "Always" | addsalt.lab == "Often", "Always/Often", addsalt.lab),
                            levels=c("Always/Often", "Sometimes", "Never/Rarely")))


## Re-order diplotype variable for tables & P-value determinations (ref = AVI/AVI)
#analysis <- analysis %>% mutate(diplos_AA = factor(diplos_AA, levels=c("AVI/AVI", "AVI/PAV", "PAV/PAV", "AVI/AAV", "AAV/PAV")))

## Demographic, behavioral & risk factors
vars_to_summarize <- c(
  age="Age, years",
  sex="Sex",
  bmi="BMI, kg/m2",
  smoke_level.lab="Smoking",
  physact_level="Physical Activity, MET/wk",
  alch_freq.lab="Alcohol Frequency",
  ALC="Alcohol intake",
  educ_level.lab="Education Level",
  income_level.lab = "Income Level",
  glu="Glucose, mmol/L",
  hba1c_max="HbA1C, %",
  sbp="SBP, mmHg",
  dbp="DBP, mmHg",
  tg="Triglyceride, mmol/L",
  ldl="LDL, mmol/L",
  hdl="HDL, mmol/L",
  coffee_QT="Coffee, cups/day",
  tea_QT="Tea, cups/day",
  addsalt_3lvl.lab="Added salt"
  )

# print summary table
t1_summary <- print_summary_table(
  vars_to_summarize = vars_to_summarize, 
  var_strata = "diplos_AA", #var_strata = "diplos_AA", 
  var_strata_order = c("AVI/AVI", "AVI/PAV", "PAV/PAV", "AVI/AAV", "AAV/PAV"),
  p_print = T, p_adjust = "agesex", digits = c(1,0,4), data=analysis) 

#t1_summary %>% write.csv("../data/processed/analysis/EUR_descr_diploAA_taste_v2.csv")
t1_summary %>% kable()


## NOTE: 0 digits for BP vars
#tapply(analysis$sbp, analysis$diplos_AA, mean_sd, d=0)
#tapply(analysis$dbp, analysis$diplos_AA, mean_sd, d=0)

```


## Check whether diplotypes captured expected phenotypic differences in bitter-related diet trait

```{r}
analysis <- analysis %>% mutate(
  taste_diplos.num = case_when(
    taste_diplos == "AVI/AVI" ~ 0,
    taste_diplos == "AVI/PAV" ~ 1,
    taste_diplos == "PAV/PAV" ~ 2,
    TRUE ~ as.numeric(NA)),
  taste_alleles = (rs713598_G+rs1726866_G+rs10246939_C),
  alch_heavydrinker = as.factor(ifelse(alch_freq.lab == "Daily or almost daily" | alch_freq.lab == "3-4 per week", 1, 0)),
  alch_lightdrinker = as.factor(ifelse(alch_freq.lab == "Special occasions only" | alch_freq.lab == "Never", 1, 0))
    )

## Compare continuous diet traits by diplotypes - P-values
rbind.data.frame(
  print_lm(exposure = "taste_diplos.num", outcome="coffee_QT", covariates = "age+sex", label="Coffee"),
  print_lm(exposure = "taste_diplos.num", outcome="tea_QT", covariates = "age+sex", label="Tea"),
  print_lm(exposure = "taste_diplos.num", outcome="raw_veg", covariates = "age+sex", label="Raw vegetables"),
  print_glm(exposure = "taste_diplos.num", outcome="alch_heavydrinker", covariates = "age+sex", label="Heavy drinker"),
  print_glm(exposure = "taste_diplos.num", outcome="alch_lightdrinker", covariates = "age+sex", label="Heavy drinker")
  ) %>%
  mutate(across(c(beta, se, p), ~as.numeric(.))) %>% mutate(across(where(is.numeric), ~round(., 3))) %>%
  kable()

```



#### Compare food choice patterns across TAS2R38 diplotypes


```{r}

dietPC.rot <- dietPCs.dat$rotation %>% 
  as.data.frame() %>%
  mutate(Diet=diet_labels[rownames(.)], .before=PC1) %>%
  arrange(PC1)

# ==================================
## Dot plot of eigenvalues (& PVE)  
# ==================================
dietPC_summary <- as.data.frame(summary(dietPCs.dat)$importance %>% t()) %>%
  mutate(dietPC=1:nrow(.),
         Eigenvalues = `Standard deviation`^2) 

p_dietPC_summary <- dietPC_summary %>% 
  ggplot(aes(x=dietPC, y=`Eigenvalues`)) +
  ggtheme +
  scale_x_continuous(breaks=seq(1,24,1), expand=c(0.01,0.5)) +
  scale_y_continuous(limits=c(0, 2.65), breaks=seq(0,2.5,0.5),
                     sec.axis = sec_axis(~.*4, name = "Percent Variance Explained (%)", breaks = seq(0,10,2))) +
  geom_point() + geom_line(size=0.15) +
  geom_hline(yintercept = 1, linetype="dashed") +
  geom_hline(yintercept = 0, linewith=0.25) +
  xlab("# Diet PCs") + 
  theme(panel.grid.major.y = element_line(),
        axis.title = element_text(face="bold"))

#p_dietPC_summary %>% ggsave(filename = "../output/plot_dietPC_summary.pdf", height=4, width=6)
p_dietPC_summary


# =====================================
## Summary plots of factor loadings
# =====================================
# Waterfall plot 
plot_dietPC_waterfall.fun(dietPC.rot)

# Heat Map 
dietPC.mat <- as.matrix(dietPC.rot[,-1])
rownames(dietPC.mat) <- dietPC.rot$Diet
#pdf("../output/plot_hm_dietPC_v2.pdf", height=4, width=12)
heatmap.2(dietPC.mat,
          col=palettes$hm_palette,
          scale = "none", 
          linecol =  "black",
          Rowv=F, Colv=F,
          colsep = seq(0,24,1), sepwidth = 0.05,
          density.info = "none", trace = "none", margins = c(5,10),
          dendrogram = "none", #key="False",
          cexRow = 0.55, cexCol = 0.75,
          offsetRow=0, offsetCol = 0) 
#dev.off()

```


#### Examine adherence to the food choice patterns by canonical TAS2R38 diplotypes

```{r}
# =======================================
# Summarize EMMs for dPCs by diplotype
# =======================================
dietPCs <- c(paste0("dietPC", 1:24))

## Filter by significance & after MC adjustment
#dietPC_pstat %>% filter(P_val_f <0.05) ; dietPC_pstat %>% filter(P_val_f <0.05/24)
#dietPC_pstat %>% filter(P_val_t <0.05) ; dietPC_pstat %>% filter(P_val_t <0.05/24)
pcs_to_plot <- c(3,4,6,11,12,13,15,16,17) # 3,6,11,12 remain after MC based on F-test


# ==========================================
## Plot all F-test statistics for dPCs
# ==========================================
plot_bar_dPCftest <- dietPC_pstat %>%
  ggplot(aes(x=DietPC, y=F_stat)) + 
  geom_bar(stat="identity", fill="darkgrey") + 
  ggtheme + 
  theme(#axis.text.x = element_blank(),
        axis.text.y=element_text(size=6), axis.title.y = element_text(size=7, face="bold"),
        axis.ticks.x = element_blank(),
        axis.text.x = element_text(size=7),
        panel.border = element_blank(),
        panel.grid.major.y = element_line()) + 
  geom_hline(yintercept = 0, linewidth=0.25) +
  ylab("F statistic") + xlab("") +
  scale_x_continuous(limits=c(0,25), breaks=seq(1,24,1), expand = c(-0.01,0)) +
  geom_vline(xintercept = 0.3, linewidth=0.25, color="black")

#plot_bar_dPCftest %>% ggsave(filename="../output/plot_dPC_Ftest_v2.pdf", height=1.5, width=8)


# ==========================================
## Plot all T-test statistics for dPCs
# ==========================================
plot_bar_dPCttest <- dietPC_pstat %>%
  ggplot(aes(x=DietPC, y=T_stat)) + 
  geom_bar(stat="identity", fill="darkgrey", width=0.8) + 
  ggtheme + 
  theme(#axis.text.x = element_blank(),
        axis.text.x = element_text(size=7), axis.text.y = element_text(size=6), 
        axis.title.x = element_text(size=7, face="bold"), axis.title.y = element_text(size=7, face="bold"),
      # panel.border = element_blank(),
        panel.grid.major.y = element_line()) + 
  geom_hline(yintercept = 0, linewidth=0.1) +
  ylab("T-statistic") + xlab("Diet Principal Components") +
  scale_x_continuous(limits=c(0,25), breaks=seq(1,24,1), expand = c(-0.01,0)) +
  scale_y_continuous(limits=c(-5.25,3.5)) #+
 # geom_vline(xintercept = 0.3, linewidth=0.25, color="black")

#plot_bar_dPCttest %>% ggsave(filename="../output/plot_dPC_Ttest_v2_labels.pdf", height=1.75, width=8)


# =====================
## Plot side-by-side
# =====================
ggarrange(plot_bar_dPCftest, plot_bar_dPCttest, nrow=2, align="hv")

```


```{r}
# ==========================================
## Plot EMM of dPCs
# ==========================================
plot_emm_dPC_signif <- dietPC_emm %>%
  filter(outcome %in% c(dietPCs[pcs_to_plot])) %>%
  mutate(outcome=factor(outcome, levels=c(paste0("dietPC", pcs_to_plot)), labels=c(paste0("Diet PC", pcs_to_plot)))) %>%
  #mutate(exp=factor(exp, levels=c("AVI/AVI", "AVI/PAV","PAV/PAV","AVI/AAV", "AAV/PAV"))) %>%
  ggplot(aes(x=exp, y=emmean, ymin=lower.CL, ymax=upper.CL, color=exp)) + 
  facet_wrap(~outcome, nrow=1) + 
  geom_hline(yintercept = 0, linewidth=0.15) +
  geom_point(size=2.25) + geom_errorbar(width=0.55) + 
  scale_color_manual(values=palettes$nature_diplosAA_canonical_yrb, name="TAS2R38 Diplotype") + 
  ylab("Marginal Mean (95% CI) PC score") + xlab("TAS2R38 Diplotype") +
  ggtheme+theme(axis.text.x = element_text(angle=35, hjust=1), legend.position = "bottom") +
    theme(axis.title = element_text(face="bold", color="black", size=8)) +
  theme(panel.grid.major.y = element_line())

#plot_emm_dPC_signif %>% ggsave(filename="../output/plot_emm_dPC_signif_v3.pdf", height=3.25, width=10)
plot_emm_dPC_signif
```


## Primary Regression Analysis
#### Asssociations of canonical TAS2R38 diplotype with markers of glucose homeostasis random glucose & a1c

For the primary analysis of TAS2R38 diplotype with random glucose, we used generalized linear models with 4 sets of covariates.

* Exposure: categorical diplotype (reference = AVI/AVI)
* Outcomes: continunous glucose traits
* Covariates: 4 models; base, bmi, lifestyle, food choices


```{r, error=F, out.width="90%"}
AAs <- c(" AVI/AVI", "  AVI/PAV", "  PAV/PAV")

## Print table of associations of Diplos with RG ---------------------------
tab_lm_diplo_rg <- read.csv("../data/processed/analysis/EUR_lm_diplo_x_rg_v2.csv") %>% 
  mutate(beta_se = sprintf("%s (%s, %s)", round(beta, 3), round(beta-1.96*se, 3), round(beta+1.96*se, 3))) %>%
  mutate(Diplo = rep(AAs, 4), .before=beta) %>%
  select(Model=model, Diplotype=Diplo, n, Beta_95CI=beta_se, P=p, Ftest_P=f_p, Ttest_P=t_p) 


## Print table of associations of Diplos with A1c ---------------------------
tab_lm_diplo_a1c <- read.csv("../data/processed/analysis/EUR_lm_diplo_x_a1c_v2.csv") %>% 
  mutate(beta_se = sprintf("%s (%s, %s)", round(beta, 3), round(beta-1.96*se, 3), round(beta+1.96*se, 3))) %>%
  mutate(Diplo = rep(AAs, 4), .before=beta) %>%
  select(Model=model, Diplotype=Diplo, n, Beta_95CI=beta_se, P=p, Ftest_P=f_p, Ttest_P=t_p)


## Combine tables & save
tab_lm_diplo_gluA1c <- left_join(tab_lm_diplo_rg, tab_lm_diplo_a1c, by=c("Model", "n", "Diplotype"), suffix = c("_RG", "_HbA1c")) %>%
  mutate(across(c("P_RG", "Ftest_P_RG", "Ttest_P_RG", "P_HbA1c", "Ftest_P_HbA1c", "Ttest_P_HbA1c"), ~round(., digits=3))) %>%
  mutate(across(c("Beta_95CI_RG", "Beta_95CI_HbA1c"), ~gsub("NA ", "Reference", gsub("[(]NA.*", "", .)))) 

#tab_lm_diplo_gluA1c %>% write.csv("../data/processed/analysis/EUR_tab_lm_diplo_x_glucA1c_v2.csv")
tab_lm_diplo_gluA1c %>%
  mutate_all(~ifelse(is.na(.), "-", .)) %>%
  kable(caption = "Association of TAS2R38 diplotype with markers of glucose homeostasis")

```


#### Asssociations of canonical TAS2R38 diplotype with markers of glucose homeostasis by fasting time 

```{r}

## Glucose stratified by fasting time in all models ---------------------------
tab_lm_diplo_gluXfast <- read.csv("../data/processed/analysis/EUR_lm_diplo_x_gluXfast_v2.csv") %>%
  mutate(beta_se = sprintf("%s (%s, %s)", round(beta, 3), round(beta-1.96*se, 3), round(beta+1.96*se, 3))) %>%
  mutate(Diplo = rep(AAs, 20), .before=beta) %>%
  select(model_fast, model, fast, Diplo, n, beta_se, p, f_p, t_p) 
#tab_lm_diplo_gluXfast %>% write.csv("../data/processed/analysis/EUR_tab_lm_rg_fastcat_allModels_v2.csv")


#tab_lm_diplo_dluXfast %>% select(-model_fast) %>%
#  pivot_wider(values_from=c(beta_se, p, f_p, t_p), names_from = model, names_glue = "{model}_{.value}", names_sort = F) %>%
#  write.csv("../data/processed/analysis/EUR_tab_lm_rg_fastcat_allModels_v2.csv")
tab_lm_diplo_gluXfast %>% select(-model_fast) %>%
  select(Model=model, Fasting=fast, Diplo, n, Beta_95CI=beta_se, P=p, Ftest_P=f_p, Ttest_P=t_p) %>%
  mutate(across(c("P", "Ftest_P", "Ttest_P"), ~round(., digits=4))) %>%
  mutate(across("Beta_95CI", ~gsub("NA ", "Reference", gsub("[(]NA.*", "", .)))) %>%
  mutate_all(~ifelse(is.na(.), "-", .)) %>%
  kable(caption = "Association of TAS2R38 diplotype with markers of glucose homeostasis by FASTING time ")


# plot of EMM glucose by fasting time in diet patterns model
emm_gluXfast_diet <- read.csv("../data/processed/analysis/EUR_emm_diplo_x_gluXfast_diet_v2.csv")

p_emm_gluXfast_diet <- emm_gluXfast_diet %>%
  mutate(fast = factor(fast, levels=c("6+hr", "5hr", "4hr", "3hr", "0to2hr"),
                       labels =c(">6 hrs", "5 hrs", "4 hrs", "3 hrs", "0-2 hrs")),
         Diplotypes = factor(exp, levels=c("AAV/PAV","PAV/PAV", "AVI/PAV", "AVI/AVI","AVI/AAV"))) %>%
  ggplot(aes(y=emmean, x=fast, ymin=lower.CL, ymax=upper.CL, color=Diplotypes)) + 
  coord_flip() +
  geom_hline(yintercept = seq(4.8,5,0.025), breaks=seq(4.825, 4.975, 0.025), linewidth=0.15, color="grey") +
  geom_point(size=1.715, position=position_dodge(0.5)) + 
  geom_errorbar(width=0.35, position=position_dodge(0.5), linewidth=0.65) + 
  scale_color_manual(values=palettes$nature_diplosAA_canonical_yrb, name="TAS2R38 Diplotype") +
  ggtheme + 
  ylab("Marginal mean (95% CI) blood glucose, mmol/L") + 
  xlab("Time since last meal, hours") + 
  theme(axis.title = element_text(face="bold", size=10),
        axis.text = element_text(size=12))

#p_emm_gluXfast_diet %>% ggsave(filename="../output/plot_emm_gluXfast_v2.pdf", height=4.5, width=5)
p_emm_gluXfast_diet

```


#### End of primary analysis tables & figures 





