---
title: "results_summary"
output: html_document
date: "2024-09-24"
---

```{r setup, include=FALSE, echo=F, error=F, warning=F, message=F}
knitr::opts_chunk$set(echo = F, error = F, warning = F, fig.align="center", 
                      out.width = "90%", out.height = "50%", fig.height=5)

#BiocManager::install("ComplexHeatmap")
library("ComplexHeatmap")

lapply(c("tidyverse", "data.table", "paletteer", "ggpubr", "ggtext", #"gplots", "R3port", 
         "xtable", "knitr", "tinytex", "emmeans"),  library, character.only = TRUE)

lapply(list.files("../../pantry/functions/", full.names = T), source)
source("../scripts/pantry.R", echo=F)
```


```{r}
###############################
##   Load data by ancestry   ##
###############################

# system arguments
EUR=c("European"="EUR")
tag="v6"

# load postprocessed dataframe
postprocessed <- readRDS("../data/processed/ukb_postprocessed_EUR_20241227.rda")

analysis <- readRDS(paste0("../data/processed/ukb_analysis_EUR.rda"))
range(analysis$glu) #: 2.116 - 8.002
  
#load dietPCs
dietPCs.dat <- readRDS(paste0("../data/processed/diet/ukb_EUR_dietPCs.rda"))
dietPC_emm <- read.csv(paste0("../data/processed/descr_tab_emm_dietPC_diplos_sexageac_", tag,".csv"))
dietPC_pstat <- read.csv(paste0("../data/processed/descr_tab_teststat_dietPC_diplos_sexageac_",tag,".csv"))

taste_diplos.labs <- c("AVI/AVI", "AVI/PAV", "PAV/PAV")
taste_diplos <- c("AVI/AVI", "AVI/PAV", "PAV/PAV")
taste_diplos_005.labs <- c(taste_diplos.labs, "AVI/AAV", "AAV/PAV")

## Add glucose values with mg/dl
analysis <- analysis %>% mutate(
  glu_mgdl = glu*18.018,
  glu2hr_mgdl = glu2hr*18.018)

```

\n
 
  
## *Inclusion/Exclusion criteria*
* Unrelated, EUR genetic ancestry, without T2D
* No diabetes data (Eastwood algorithm + HbA1c <5.7%)
* Complete data for glucose, genetic ancestry, and behavioral/SES covariates


```{r incl-anc}

prop.table(table(postprocessed$diplos_common_AA)) %>% t() %>% 
  kable(caption="All diplotypes wth Freq > 0.001 in the TOTAL EUR sample, before sample exclusions (N=3612209)")

postprocessed <- postprocessed %>% 
  mutate(n_glu_range=ifelse(glu>=2.116 & glu <=8.002,1,0)) %>%
  mutate(n_glu_outliers=find_outliers.fun(glu, recode_outliers = "T")) %>%
  mutate(n_fasting_gt24=ifelse(fasting_hrs > 24, 1,0)) %>%
  mutate(n_include=ifelse(n_geno==1 & n_t2d_contrl==1 & n_complete==1 & 
                            n_fast_24hr==1 & n_glu_range==1,1,0)) %>%
  mutate(n_include_descr=ifelse(n_include==1 & diplos_common_AA %in% taste_diplos_005.labs,1,0),
         n_include_analysis=ifelse(n_include==1 & diplos_common_AA %in% taste_diplos.labs,1,0))

postprocessed %>%
  reframe("European, *n* (%)"=n_pct(N_ancestry, level="1"),
          "Genotyping, *n* (%)"=n_pct(n_geno, level="1"),
          "   Exclude if missing genotype data, *n* (%)"=n_pct(n_geno, level="0"),
          "T2D Controls, *n* (%)"=n_pct(n_t2d_contrl, level="1"),
          "   Exclude T2D cases, *n* (%)"=n_pct(t2d_case.f, level="Case"),
          "   Exclude outliers for Gluose, n(%)"=n_pct(n_glu_outliers, level="T"),
          "Complete covariate data, *n* (%)"=n_pct(n_complete, level="1"),
          "   Exclude if missing covariate data, *n* (%)"=n_pct(n_complete, level="0"),
          "Fasting ≤24 hr, *n* (%)"=n_pct(n_fast_24hr, level="1"),
          "   Exclude if fasting >24 hr, *n* (%)"=n_pct(n_fasting_gt24, level="1"),
          "Available & complete data, *n* (%)"=n_pct(n_include, level="1"),
          "   Total excluded, *n* (%)"=n_pct(n_include, level="0"),
          "Descriptive sample (*TAS2R38* diplo >0.005), *n* (%)"=n_pct(n_include_descr, level="1"),
          "Analytical sample (*TAS2R38* taste diplo), *N* (%)"=n_pct(n_include_analysis, level="1")
          ) %>% 
  t() %>%
  kable(caption = "Inclusion/Exclusion criteria & sample breakdown")

## Table of diplotypes in complete/available data sample
print_summary_table(
  data=postprocessed %>% 
    filter(n_include==1), vars_to_summarize = c("diplos_common_AA"="diplos_common_AA")) %>%
  kable()

## Table of diplotypes across study centers
as.data.frame(read.csv(paste0("../data/processed/descr_tab_tastediplo_ac_",tag,".csv")) %>%
  t())[,-2] %>% 
  mutate(P_Chi.Sq = c(chisq.test(analysis$ac.f, analysis$taste_diplos)$p.value, rep(NA, nrow(.)-1))) %>%
  kable()

```

## Distribution of TAS2R38 Haplotypes & Diplotypes 
We examined the distributions of common (frequency >0.0005) TAS2R38 genotypess 
```{r, error=F, message=F, fig.height==4, fig.width=6}
# ================================
## Summary of haplotypes
# ================================

## haplo_legend
pal_haplo_legend <- c(palettes$nature_haplos_dominant)[c(1,4,3)]
names(pal_haplo_legend) <- paste(c("Nontaster", "Taster", "Non-canonical"), "haplotype")

tab_haplos_AA <- analysis %>%
  select(id, haplo_0_aa, haplo_1_aa) %>%
  pivot_longer(-id, values_to="haplo") %>%
  group_by(haplo) %>% reframe(n=n(), pct=(n()/nrow(.))*100) %>%
  arrange(n) %>% mutate(haplo_order = factor(haplo, levels=haplo[order(n, decreasing = T)])) %>%
  mutate(diplo_legend = factor(ifelse(haplo_order == "AVI", "Nontaster haplotype", 
                                      ifelse(haplo_order == "PAV", "Taster haplotype", "Non-canonical haplotype")), 
                               levels=paste(c("Nontaster", "Taster", "Non-canonical"), "haplotype"))) %>%
  #Filter to "common" haplotypes (>1% of the cohort; >0.005, frequency)
  filter(pct>0.05) ; plot_haplos_AA <- tab_haplos_AA %>% 
  ggplot(aes(x=haplo_order, y=pct, fill=diplo_legend)) + 
  geom_bar(stat = "identity") + 
  scale_x_discrete(name = "**_TAS2R38_ Haplotype**") + 
  scale_y_continuous(limits=c(0,62), breaks=seq(0,60,10)) +
  ylab("Frequency (%)") + 
  scale_fill_manual(values = pal_haplo_legend, name=expression(italic("TAS2R38") ~ "Haplotypes (n)")) + 
  ggtheme_fancy + 
  theme(axis.text.x = element_text(size=8, color = "black", face="italic"), #angle=25, hjust=1, 
        axis.title.x = element_markdown(),
        axis.title = element_text(color = "black", face = "bold", size=10),
        axis.text.y = element_text(size=8, color = "black"),
        
        legend.position = "inside",
        legend.text = element_text(size=8), legend.title=element_blank(),
        legend.position.inside = c(0.99, 0.98),  # Moves legend to top-right inside plot
        legend.justification.inside = c(1, 1),  # Aligns legend's top-right corner
        legend.background = element_rect(fill = "white", color = "black", linewidth=0.15), 
          
        legend.key.size = unit(1, 'line'),
        legend.margin = margin(3,3,3,3, unit="pt")
        ) +
  
    geom_text(aes(label=n), vjust=-0.35, size=3.0)

plot_haplos_AA
#plot_haplos_AA %>% ggsave(filename=paste0("../output/descr_plot_haploAA005_",tag,".pdf"), height = 4, width=4.5)
```

```{r}
# ================================
## Summary of Diplotypes
# ================================
## Create variable in analysis for **common diplotypes**
diplos_gt005_AA <- names(which(prop.table(table(analysis$diplos_common_AA))>0.005))
analysis <- analysis %>% mutate(
  diplos_AA = factor(ifelse(diplos_common_AA %in% diplos_gt005_AA, diplos_common_AA, NA), levels=c(diplos_gt005_AA)))

## diplo_legend
pal_diplo_legend <- c(palettes$nature_diplos_dominant)[c(1,3:5)]
names(pal_diplo_legend) <- paste(c("Nontaster", "Taster", "Supertaster", "Non-canonical"), "diplotype")

## Table
tab_diplos_AA <- analysis %>% 
  filter(complete.cases(diplos_AA)) %>% #diplos_common_AA %in% diplos_common_AA) %>% 
  reframe(n=table(diplos_AA), pct=prop.table(table(diplos_AA)), diplos_AA=names(table(diplos_AA))) %>%
  arrange(-n) %>% 
  mutate(color = factor(ifelse(diplos_AA == "AVI/AVI", "Nontaster",
                                 ifelse(diplos_AA == "AVI/PAV", "Taster",
                                        ifelse(diplos_AA == "PAV/PAV", "Supertaster", "Non-canonical"))),
                          levels=c("Nontaster","Taster","Supertaster","Non-canonical"))) %>%
  mutate(diplos_AA_order=factor(diplos_AA, levels=c("AVI/AVI", "AVI/PAV", "PAV/PAV", "AVI/AAV", "AAV/PAV")),
         diplo_legend = factor(paste(color, "diplotype"), levels=
                                 paste(c("Nontaster","Taster","Supertaster","Non-canonical"), "diplotype"))) %>%
  filter(pct>0.005) ; plot_diplo_AA <- tab_diplos_AA %>%
  ggplot(aes(x=diplos_AA_order, y=pct*100, fill=diplo_legend)) + 
  geom_bar(stat = "identity") + scale_x_discrete(name = "**_TAS2R38_ Diplotypes**") + 
  scale_y_continuous(limits=c(0,52), breaks=seq(0,50,10)) + ylab("Frequency (%)") + 
  scale_fill_manual(values = pal_diplo_legend, name=expression(italic("TAS2R38") ~ "Diplotypes (n)")) + 
  ggtheme_fancy +
    theme(axis.text.x = element_text(size=8, color = "black", face="italic"), #angle=25, hjust=1, 
          axis.text.y = element_text(size=8, color = "black"),
          axis.title.x = element_markdown(),
          axis.title = element_text(color = "black", face = "bold", size=10),
          
          legend.position = "inside",
          legend.text = element_text(size=8), legend.title=element_blank(),
          legend.position.inside = c(0.99, 0.98),  # Moves legend to top-right inside plot
          legend.justification.inside = c(1, 1),  # Aligns legend's top-right corner
          legend.background = element_rect(fill = "white", color = "black", linewidth=0.15), 
          
          legend.key.size = unit(1, 'line'),
          legend.margin = margin(3,3,3,3, unit="pt"),
          
          ) +  # Optional: Add background
  
  geom_text(aes(label=n), vjust=-0.35, size=3.0)

plot_diplo_AA #%>% ggsave(filename=paste0("../output/descr_plot_diploAA005_order2_",tag,".pdf"), height = 4, width=6)


# =================================================
# Panel plot of TAS2R38 haplotypes & diplotypes 
# =================================================
plot_haplo_diplo_AA <- ggarrange(plot_haplos_AA, "",  plot_diplo_AA, ncol=3, widths=c(0.75, 0.05, 1), align = c("hv"), labels=c("A", "", "B")#,
          #common.legend = T, legend = "none"
          )

plot_haplo_diplo_AA
plot_haplo_diplo_AA %>% ggsave(filename=paste0("../output/manuscript/descr_plot_genoAA005_panel_",tag,"_vManuscript.pdf"), height=3.75, width=9.15)
 
```


## Descriptive characteristics 
Next, we summarized demographic, behavioral and health characteristics among all identified TAS2R38 diplotypes. P-values provided are for T-tests (of any differenes among CANONICAL diplotypes) 
and for P-trend, across additive diplotypes (0,1,2 PAV haplotypes)

```{r, include=T}
read.csv(paste0("../data/processed/descr_tab_diplo_AA_taste_",tag,".csv")) %>% kable(caption="Descriptive Table 1")

## Create analytical dataframe with only canonical diplotypes (N=218688)
#analysis_all <- analysis
analysis <- analysis %>% filter(!is.na(taste_diplos))

## Pairwise tests
analysis <- analysis %>% mutate(
  frequent_alch_intake = factor(ifelse(alch_freq.lab %in% c("Daily or almost daily", "3-4 per week"), 1, 0))) %>%
  mutate(income_gt50k = factor(ifelse(income_level.lab %in% c("from_52000_to_100000", "gt_100000"), 1, 0))) %>%
  mutate(taste_diplos_pav = relevel(taste_diplos, ref="PAV/PAV"),
         taste_diplos_pavavi = relevel(taste_diplos, ref="AVI/PAV"))

## Heavy drinking pairwise comparisons
#summary(glm(alch_heavydrinker ~ taste_diplos_pav + sex+ age, family=binomial("logit"), data=analysis))

## Pairwise comparisons vs. PAV/PAV 
addn_comparisons <- c(frequent_alch_intake ="≥3 drinks/week",alch_drinks_per_week="Drinks/week", 
                      cigarettes_per_day="Cigarettes/day", income_gt50k="Income ≥ 50k", 
                      coffee_QT="coffee", raw_veg_QT="raw_veg", tea_QT="tea")
lapply(c("Female", "Male"), function(x) {
  print_summary_table(vars_to_summarize = addn_comparisons, var_strata = "taste_diplos", 
                      p_adjust = "age", var_strata_order = taste_diplos, 
                      data=analysis %>% filter(sex == x)) 
  }) %>% kable(caption = paste0("Additional comparisons for Table 1 Females (top) & Males (bottom)")) 

#lapply(names(addn_comparisons), function(v) {
#  do.call(rbind.data.frame, lapply(c("taste_diplos", "taste_diplos_pav", "taste_diplos_pavavi"), function(d) {
#    if(is.numeric( (analysis %>% select(v)) [[1]] ) ) {
#      print_lm(exposure=d, outcome=v, covariates = "age+sex", label = addn_comparisons[v][[1]]) } else {
#    print_glm(exposure=d, outcome=v, covariates = "age+sex", label = addn_comparisons[v][[1]]) }
#    }) )
#  })

## Additinal p-values for Unadjusted chisquare or p-trends for Table 1 *********
table1_pvals <- do.call(rbind.data.frame, lapply(1:length(vars_to_summarize), function(v) {
  dat <- analysis %>% select(taste_diplos.num, var=names(vars_to_summarize[v]))
  if(is.numeric(dat$var) ==T) {
    cbind.data.frame(Var=vars_to_summarize[[v]], Pvalue=summary(lm(var~taste_diplos.num, data=dat))$coef[2,4])
  } else {
    cbind.data.frame(Var=vars_to_summarize[[v]], Pvalue=chisq.test(dat$taste_diplos.num, dat$var)$p.value)
  }
})) %>% mutate(P_formatted = format_p.fun(Pvalue))
      
table1_pvals %>% write.csv("../output/manuscript/table1_pvalues.csv")
# NOTE: P-value for added salt is wrong...

```

median [IQR]
```{r, include=F}
analysis %>% 
  filter(complete.cases(taste_diplos)) %>%
  reframe("Age, years"=median_IQR(age, d=1),
          "BMI, kg/m2"=median_IQR(bmi, d=1),
          "Cigarettes per day"=median_IQR(cigarettes_per_day, d=1),
          "PA level"=median_IQR(physact_met_excess, d=1),
          "Drinks per week"=median_IQR(alch_drinks_per_week, d=1))

analysis %>% group_by(taste_diplos) %>%
   filter(complete.cases(taste_diplos)) %>%
  reframe("Age, years"=median_IQR(age, d=1),
          "BMI, kg/m2"=median_IQR(bmi, d=1),
          "Cigarettes per day"=median_IQR(cigarettes_per_day, d=1),
          "PA level"=median_IQR(physact_met_excess, d=1),
          "Drinks per week"=median_IQR(alch_drinks_per_week, d=1))

```


#### Description of glucose levels by fasting time 
```{r}
# Check variable with delineated 6to24hr 
analysis <- analysis %>% mutate(fast_cat2 = factor(ifelse(fasting_hrs >= 6 & fasting_hrs < 12,"6to12hr", 
                                                   ifelse(fasting_hrs >= 12, "12+hr", fast_cat)),
                                                   levels=c("0to2hr", "3hr", "4hr", "5hr", "6to12hr", "12+hr"))) 

rbind.data.frame(
  print_summary_table(data=analysis %>% filter(complete.cases(taste_diplos)), vars_to_summarize = c(
  taste_diplos="Taste Diplotypes", glu_mgdl = "Random glucose, mg/dL"), digits = c(1,1),
  var_strata = "fast_cat2", var_strata_order = c("0to2hr", "3hr", "4hr", "5hr", "6to12hr", "12+hr")) %>% 
    t() %>% as.data.frame() %>% mutate(Fast=rownames(.)), 
    print_summary_table(data=analysis %>% filter(complete.cases(taste_diplos)), vars_to_summarize = c(
      taste_diplos="Taste Diplotypes", glu_mgdl = "Random glucose, mg/dL"), digits=c(1,2),
      var_strata = "fasting_hrs", var_strata_order = c(0:24)) %>% 
    t() %>% as.data.frame() %>% mutate(Fast=as.numeric(rownames(.))) %>% arrange(Fast)  ) %>%
  filter(rownames(.) %in% c("P_value", "P_trend")==F) %>%
  write.csv("../data/processed/descr_tab_gluxfasting_combined_v6.csv")

analysis %>% group_by(fasting_hrs) %>%
  summarise(n=n(), mean=mean(glu_mgdl),
            sd=sd(glu_mgdl)) %>%
  mutate(fast_cat=factor(c(rep("0to2hr",3), "3hr", "4hr", "5hr", rep("6to12hr",6), rep("12+hr",13)),
                         levels=c("0to2hr", "3hr", "4hr", "5hr", "6to12hr", "12+hr"))) %>%
  ggplot(aes(x=as.factor(fasting_hrs), y=mean-80, ymin=(mean)-80, ymax=(mean+1.96*sd)-80)) + 
  geom_bar(stat="identity") + 
  geom_errorbar() + 
  facet_grid(~fast_cat, scales="free_x", space="free_x")

```

## TAS2R38 and food choices 
##### Check whether diplotypes captured expected phenotypic differences in bitter-related diet trait

```{r}

## Compare continuous diet traits by diplotypes - P-values
bitter_foods <- c(
  coffee_QT="Coffee, cups/day", tea_QT="Tea, cups/day",
  raw_veg="Raw vegetables, tbsp/day", alch_drinks_per_week="Alcohol, drinks per week",
  alch_heavydrinker="Heavy drinker"
)
  
## save summary table for taste_diplos, only for p-values
t1_bitterfood <- print_summary_table(
  vars_to_summarize = bitter_foods, 
  var_strata = "taste_diplos_pav", #var_strata = "diplos_AA", 
  var_strata_order = c("AVI/AVI", "AVI/PAV", "PAV/PAV"),
  p_types = c("descriptive", "trend"),
  p_print = T, p_adjust = "agesex", digits = c(2,1,6), data=analysis) 

t1_bitterfood %>% kable()
```

##### Compare food choice patterns across TAS2R38 diplotypes

```{r, fig.width=6, fig.height=4, warning=F}

dietPC.rot <- dietPCs.dat$rotation %>% 
  as.data.frame() %>%
  mutate(Diet=diet_labels_descriptive[rownames(.)], .before=PC1) %>%
  arrange(PC1)

# ==================================
## Dot plot of eigenvalues (& PVE)  
# ==================================
dietPC_summary <- as.data.frame(summary(dietPCs.dat)$importance %>% t()) %>%
  mutate(dietPC=1:nrow(.),
         Eigenvalues = `Standard deviation`^2) 

p_dietPC_summary <- dietPC_summary %>% 
  ggplot(aes(x=dietPC, y=`Eigenvalues`)) +
  ggtheme_fancy +
  scale_x_continuous(breaks=seq(1,24,1), expand=c(0.01,0.5)) +
  scale_y_continuous(limits=c(0, 2.65), breaks=seq(0,2.5,0.5),
                     sec.axis = sec_axis(~.*4, name = "Percent Variance Explained (%)", breaks = seq(0,10,1))) +
  geom_point() + geom_line(size=0.15) +
  geom_hline(yintercept = 1, linetype="dashed") +
  geom_hline(yintercept = 0, linewith=0.25) +
  xlab("# Diet PCs") + 
  theme(panel.grid.major.y = element_line(), panel.grid.minor.y = element_line(),
        axis.title = element_text(face="bold"))

p_dietPC_summary %>% ggsave(filename = paste0("../output/suppl_plot_dietPCeigenvals_",tag,".pdf"), height=4, width=6)
p_dietPC_summary
```


#### Examine adherence to the food choice patterns by canonical TAS2R38 diplotypes
```{r, warning=F}
# =====================================
## Summary plots of factor loadings
# =====================================
dietPC.rot <- dietPCs.dat$rotation %>% 
  as.data.frame() %>%
  mutate(Diet=diet_labels_descriptive[rownames(.)], .before=PC1) %>%
  arrange(PC1)
dietPC.mat <- as.matrix(dietPC.rot[,-1])
rownames(dietPC.mat) <- dietPC.rot$Diet

## Filter by significance & after MC adjustment
pcs_to_plot <- c(3,4,6,11,12,13,15,17) # 311,12 remain after MC based on F-test

colnames(dietPC.mat) <- gsub("PC", "", colnames(dietPC.mat))
#pdf(paste0("../output/plot_res_hm_dietPCs_annotated_",tag,"_4x7.pdf"), height=3.75, width=9)
hm_diplo_dietPC <- Heatmap(dietPC.mat, col=palettes$hm_palette_rwb, 
        cluster_columns = F, cluster_rows = F, 
        column_title = "Diet Principal Components", column_title_side = "bottom",
        column_labels = paste0("PC",1:24),
        column_title_gp = gpar(fontface="bold", fontsize=8),
        row_title = " ", #row_title = "Diet Traits", 
        row_title_gp = gpar(fontface="bold", fontsize=7), 
        row_names_side="left", row_names_gp = gpar(fontsize = 7), 
        column_gap = unit(0.5, "mm"), column_split=seq(1,24,1),
        column_names_gp = gpar(fontsize=8), 
        column_names_rot = 90, column_names_centered = T,
        width=unit(12.5, "cm"), height=unit(7, "cm"),
        heatmap_legend_param = list(
          at = c(seq(-1,1,0.5)), labels = seq(-1,1,0.5), 
          title = "Rotated PC\nFactor Loadings", 
          legend_height = unit(3.5, "cm"), legend_width=unit(2,"cm"),
          grid_width=unit(0.35,"cm"), labels_gp = gpar(fontsize=8),
          title_position = "topleft"),
        # Add * annotations for substantial loadings for significantly associated PCs
        cell_fun = function(j, i, x, y, width, height, fill) {
        if(abs(dietPC.mat[i, j]) > 0.3 & colnames(dietPC.mat)[j] %in% c(pcs_to_plot)) #c(paste0("PC",pcs_to_plot)))
          grid.text(sprintf("+", dietPC.mat[i, j]), x, y, gp = gpar(fontsize = 6, fontface="bold")) 
          }
        )
#dev.off()

# Notes on versions
## v1 - PC1..PC24 with vertical orientation :  column_gap = unit(0.5, "mm"), column_split=seq(1,24,1), column_names_gp = gpar(fontsize=8), 
## v2 - 1..24 with diet principal components labeled
### colnames(dietPC.mat) <- gsub("PC", "", colnames(dietPC.mat)) ; column_names_rot = 0 ; tag="v4.2"

```

```{r,  warning=F}
## Correlations of TAS2R38 haplotypes with diet traits (adjusted for sex, age, & assessment center)
lm_dietTraits <- fread("../data/processed/descr_tab_teststat_dietTraits_diplos_sexageac_v6.csv")

## Heat Map of diet traits
diplo_diet.df <- lm_dietTraits %>% select(DietTraits=Diet, Beta, SE, T_stat, P=P_val_t) %>% 
  mutate(Beta_Std=Beta/SE) %>%
  mutate(Diet=diet_labels_descriptive, .before="Beta") %>%
  as.data.frame() %>%
  left_join(dietPC.rot %>% select(Diet, PC1), by = "Diet") %>%
  arrange(PC1) 
rownames(diplo_diet.df) <- diplo_diet.df$Diet
diplo_diet.mat <- as.matrix(diplo_diet.df[,-c(1:2)])

hm_corr_diet <- Heatmap(
  diplo_diet.mat[,5], col=palettes$hm_palette_gwp_custom, 
  cluster_columns = F, cluster_rows = F,
  column_labels = "\n\nTAS2R38\nHaplotypes", column_title_side = "bottom",
  column_names_rot = 0, 
  column_names_centered = T,
  column_names_gp = gpar(fontsize=8, fontface="bold"), 
  row_names_side="left", row_names_gp = gpar(fontsize = 8), 
  width=unit(0.5, "cm"), height=unit(7.5, "cm"),
   heatmap_legend_param = list(
          #at = c(seq(-.04,.04,0.04)), labels = seq(-.04,.04,0.04), 
          at = c(seq(-5,5,2.5)), labels = seq(-5,5,2.5), 
          title = "Beta\ncoefficient", 
          legend_height = unit(2, "cm"), legend_width=unit(2,"cm"),
          grid_width=unit(0.35,"cm"), 
          labels_gp = gpar(fontsize=8),
          title_position = "topleft"),
)

pdf(paste0("../output/plot_res_hm_dietPCs_annotated_",tag,"_dietTraits_4x7_v10.pdf"), height=3.55, width=9.25)
draw(hm_corr_diet+hm_diplo_dietPC, ht_gap=unit(c(0.15),"mm"))
dev.off()
```



```{r}
# ==========================================
## Plot all T-test statistics for dPCs
# ==========================================
dietPCs <- c(paste0("dietPC", 1:24))
dietPC_pstat_annotation <- dietPC_pstat %>% 
  mutate(signif = ifelse(P_val_t < 0.05/24, "**", 
                  ifelse(P_val_t >= 0.05/24 & P_val_t <0.05, "*", ""))
  )

plot_bar_dPCttest <- dietPC_pstat %>%
  ggplot(aes(x=rev(DietPC), y=T_stat)) + 
  geom_bar(stat="identity", fill="darkgrey", width=0.8) + 
  theme(
        axis.text.x = element_text(size=7, color="black"), axis.title.x = element_text(size=7, face="bold"), 
        axis.text.y = element_text(size=6, color="black"), axis.title.y = element_text(size=9, face="bold"),
        panel.grid.major.y = element_line(),  panel.background = element_rect(fill="#00000010")) + 
  geom_hline(yintercept = 0, linewidth=0.25) +
  ylab("T-statistic from linear models of\nTAS2R38 diplotype (additive) with diet PCs") + xlab("Diet Principal Components") +
  scale_x_continuous(limits=c(0,25), breaks=seq(1,24,1), expand = c(-0.01,0), labels=rev(seq(1,24,1))) +
  scale_y_continuous(limits=c(-5,5)) + 
  annotate("text", x=rev(dietPC_pstat$DietPC), y=dietPC_pstat$T_stat*1.04, vjust=0.8, label=dietPC_pstat_annotation$signif) + #c(-5.25,3.5)
  coord_flip()

plot_bar_dPCttest
plot_bar_dPCttest %>% ggsave(filename=paste0("../output/plot_res_dPCttest_grey_annotated_", tag,".pdf"), width=2.5, height=3.75)
##v6: updated data to match dietPCs derived in v6
##v2: lables 1..24; Diet Principal Components; shorter; tag="v4.2" 
#v1: dietPC1..24:  width=2.35, height=5.35
```

```{r}
# ==========================================
## Plot EMM of dPCs
# ==========================================
plot_emm_dPC_signif <- dietPC_emm %>%
  filter(outcome %in% c(dietPCs[pcs_to_plot])) %>%
  mutate(outcome=factor(outcome, levels=c(paste0("dietPC", pcs_to_plot)), labels=c(paste0("PC", pcs_to_plot)))) %>%
  #mutate(exp=factor(exp, levels=c("AVI/AVI", "AVI/PAV","PAV/PAV","AVI/AAV", "AAV/PAV"))) %>%
  ggplot(aes(x=level, y=emmean, ymin=upCI, ymax=lowCI, color=level)) + 
  ggtheme_basic+
  facet_wrap(~outcome, nrow=1, strip.position = "bottom") + 
  geom_hline(yintercept = 0, linewidth=0.3) +
  geom_point(size=2.55) + geom_errorbar(linewidth=0.65, width=0.55) + 
  scale_color_manual(values=palettes$nature_diplos_dominant, name="TAS2R38 Diplotype") + 
  ylab("Marginal Mean (95% CI)\n Diet PC score") + 
  xlab(" ") + scale_y_continuous(limits=c(-0.052,0.052), breaks=seq(-0.05, 0.05, 0.025)) +
  theme(legend.position = "top", 
        axis.title.y = element_text(color="black", size=9, face="bold"),
        axis.text.y=element_text(size=7),
        axis.text.x = element_blank(), axis.ticks.x=element_blank(),
        panel.grid.major.y = element_line(), 
        strip.background = element_blank(), 
        panel.background = element_rect(fill="#00000010"),
        strip.text = element_text(size=8, face="plain"))

plot_emm_dPC_signif %>% ggsave(filename=paste0("../output/res_plot_emm_dPCxdiplo_grey_",tag,".pdf"), height=3, width=10)
plot_emm_dPC_signif
#v2: addded stars, changed column names
#v1: height=3.5, width=8

plot_emm_dPC_signif_v2 <- dietPC_emm %>%
  filter(outcome %in% c(dietPCs[pcs_to_plot])) %>%
  mutate(outcome=factor(outcome, levels=c(paste0("dietPC", pcs_to_plot)), labels=c(paste0("PC", pcs_to_plot)))) %>%
  #mutate(exp=factor(exp, levels=c("AVI/AVI", "AVI/PAV","PAV/PAV","AVI/AAV", "AAV/PAV"))) %>%
  ggplot(aes(x=level, y=emmean, ymin=upCI, ymax=lowCI, color=level)) + 
  ggtheme_basic +
  theme_bw()+
  facet_wrap(~outcome, nrow=1, strip.position = "bottom") + 
  geom_hline(yintercept = 0, linewidth=0.3) +
  geom_point(size=2.55) + geom_errorbar(linewidth=0.65, width=0.55) + 
  scale_color_manual(values=palettes$nature_diplos_dominant, name="TAS2R38\nDiplotype") + 
  ylab("Marginal Mean (95% CI)\n Diet PC score") + 
  xlab(" ") + scale_y_continuous(limits=c(-0.052,0.052), breaks=seq(-0.05, 0.05, 0.025)) +
  theme(legend.position = "right",
        legend.title = element_text(face="bold", size=9),
        axis.title.y = element_text(color="black", size=10, face="bold"),
        axis.text.y=element_text(size=9),
        axis.text.x = element_blank(), axis.ticks.x=element_blank(),
        panel.grid.major.y = element_line(), 
        strip.background = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.background = element_rect(fill="#00000000"),
        strip.text = element_text(size=10, face="plain"))

plot_emm_dPC_signif_v2 %>% ggsave(filename=paste0("../output/res_plot_emm_dPCxdiplo_wh_",tag,"_v10.pdf"), height=2.5, width=10) 

```

#### Run between-group comparisons for PCs with significant F-tests
```{r}
analysis %>% group_by(sex, taste_diplos) %>%
  reframe(dietPC3=mean_sd(dietPC3), dietPC11=mean_sd(dietPC11), dietPC12=mean_sd(dietPC12)) %>%
  kable() ; make_pretty_lm(do.call(rbind.data.frame, lapply(c("dietPC3", "dietPC11", "dietPC12"), function(d) {
  print_lm(exposure = "taste_diplos", outcome = d, covariates = "age+sex",
           label=d, lm_trend = T) }))) %>% kable()
```


Secondary tests
```{r, include=F}
dietpc_follow_up <- c(dietPC11="dietPC11", dietPC12="dietPC12", tea_QT="Tea Intake", coffee_QT="Coffee intake", 
    milk_type.lab="Milk Type", milk_type_full_vs_low_or_nonfat_BIN="Prefer Full>Low/No-Fat Milk",
    coffee_type_decaf_vs_regular="Decaf>Caf Coffee",
    addsalt_always_often_vs_nrs="Add Salt Often/Always > Never/Sometimes", addsalt_3lvl.lab="Add Salt Frequency",
    spread_type_butter_vs_any_other="Prefer Butter > Other Spreads", 
    cheese_QT="Cheese Intake", dried_fruit_QT="Dried Fruit Intake")

# Summarise by TAS2R38 diplotype
cbind(print_summary_table(vars_to_summarize = dietpc_follow_up, digits=c(2,0,10),
    var_strata = "taste_diplos", var_strata_order = taste_diplos.labs, 
    factor_vars = c("milk_type_full_vs_low_or_nonfat_BIN", "milk_type.lab", "coffee_type_decaf_vs_regular",
                    "spread_type_butter_vs_any_other"), p_adjust = "age", data=analysis),
    print_summary_table(vars_to_summarize = dietpc_follow_up, digits=c(2,0,10),
                        var_strata = "sex", var_strata_order = c("Female", "Male"), 
    factor_vars = c("milk_type_full_vs_low_or_nonfat_BIN", "milk_type.lab", "coffee_type_decaf_vs_regular",
                    "spread_type_butter_vs_any_other"), p_adjust = "age", data=analysis) 
) %>% kable()

# summarise by TAS2R38 diplotype, stratified by sex
diet_by_diplo_sex <- do.call(rbind.data.frame, lapply(c("Female", "Male"), function(x) {
  print_summary_table(vars_to_summarize = dietpc_follow_up, digits=c(2,0,10),
    var_strata = "taste_diplos", var_strata_order = taste_diplos.labs, 
    factor_vars = c("milk_type_full_vs_low_or_nonfat_BIN", "milk_type.lab", "coffee_type_decaf_vs_regular",
                    "spread_type_butter_vs_any_other"), p_adjust = "age", data=analysis %>%
      filter(sex==x)) %>% mutate(Sex=x, .before="Total")
})) 

# summarise by sex, stratified TAS2R38 diplotype
#do.call(rbind.data.frame, lapply(taste_diplos.labs, function(x) {
#  print_summary_table(vars_to_summarize = dietpc_follow_up, digits=c(2,0,10),
#    var_strata = "sex", var_strata_order = c("Female", "Male"), 
#    factor_vars = c("milk_type_full_vs_low_or_nonfat_BIN", "milk_type.lab", "coffee_type_decaf_vs_regular",
#                    "spread_type_butter_vs_any_other"), p_adjust = "age", data=analysis %>%
#      filter(taste_diplos==x)) %>% mutate(Diplos=x, .before="Total")
#})) %>% kable()

## Summarize bitter-tasting foods
bitter_by_diplo <- do.call(rbind, lapply(names(bitter_foods[1:4]), function(f) { 
  analysis %>% rename(food=f) %>% 
    group_by(taste_diplos) %>% 
    filter(!is.na(taste_diplos)) %>%
    reframe(m=mean(food, na.rm=T), se=sd(food, na.rm=T)/sqrt(n())) %>%
    mutate(food=bitter_foods[f][[1]])
  }) ) ; bitter_by_diplo %>% 
  mutate(m_adj = ifelse(food == "Coffee, cups/day", m-1.5, 
                              ifelse(food == "Tea, cups/day", m-3, 
                                     ifelse(food == "Raw vegetables, tbsp/day", m-2, m-8)))) %>%
  ggplot(aes(x=food, y=m_adj, ymin=m_adj-1.96*se, ymax=m_adj+1.96*se, fill=taste_diplos)) +
  geom_bar(stat="identity", position=position_dodge(0.9)) + 
  facet_wrap(~food, scale="free", nrow=1) + 
  geom_errorbar(width=0.25, position=position_dodge(0.9)) + 
  scale_fill_manual(values=palettes$nature_diplos_dominant, name="TAS2R38 Diplotype") + 
  ggtheme_basic

```
**


```{r, include=F}
#### Look further into sex-stratified comparison for dietPC11 & dietPC12, which seem to be sex-specific
do.call(rbind.data.frame, lapply(c("Female", "Male"), function(i) {
  print_summary_table(vars_to_summarize = c(dietPC11=paste0(i, ": Diet PC11"), dietPC12 = paste0(i, ": Diet PC12")),
                    var_strata = "taste_diplos", var_strata_order = c("AVI/AVI", "AVI/PAV", "PAV/PAV"), 
                    data=analysis %>% filter(sex == i), p_adjust = "age") 
})) %>% kable(caption="Diet PC11 & PC12 by TAS2R38 diplotype & sex")

bind_rows(print_lm_interaction(exposure = "taste_diplos", interaction = "sex", outcome = "dietPC11",
                               model="age+sex", label_model = "Base", label_interaction = "Diet PC 11 ~ DiploxSex", 
                               print_interaction_term = T, data=analysis),
          print_lm_interaction(exposure = "taste_diplos", interaction = "sex", outcome = "dietPC12", model="age+sex", 
                     label_model = "Base", label_interaction = "Diet PC 12 ~ DiploxSex", print_interaction_term = T, data=analysis)
) %>% kable()
```
*Note: I looked at interactions of taste_diplo x sex on dietPC11 and dietPC12 which seem to be sex specific. While there were no sigificant interactions its interesting that the inheritance sems to differ for males vs. females:
* For males, AVI/AVI ~ AVI/PAV ≠ PAV/PAV (recessive coding, where both PAV haplotypes are needed to show an effect)
* For females, AVI/AVI ≠ AVI/PAV ~ PAV/PAV (dominant coding, where one PAV haplotype is needed to show an effect)


***

## Primary Analysis
#### Asssociations of canonical TAS2R38 diplotype with markers of glucose homeostasis random glucose & a1c
Descriptive comparisons: Mean (SD) with sex- and age-adjusted P-values

```{r}
fast_cat <- c("0to2hr", "3hr", "4hr", "5hr", "6+hr")
fast_cat.l <- as.list(fast_cat)

read.csv(paste0("../data/processed/descr_tab_allglycemic_tastediplos_noadj_mgdl_",tag,".csv")) %>% 
  kable(caption="Description of glycemic traits (,g/dL) with sex/age adjustment")

make_pretty_lm(read.csv(paste0("../data/processed/descr_tab_lm_allglycemic_tastediplos_sexage_mgdl_",tag,".csv"))) %>% kable()
make_pretty_lm(read.csv(paste0("../data/processed/descr_tab_lm_allglycemic_tastediplos_add_sexage_mgdl_",tag,".csv"))) %>% kable()

cat("Note: Bonferroni-adjusted P-value threshold for 3-related outcomes = P<0.05/3, 0.01667")  
```

Pairwise comparisons among all diplotypes
```{r}
outcomes.l =list(glu_mgdl="Glucose", glu2hr_mgdl="02-hr Glucose", hba1c="HbA1c")

lapply(c("glu_mgdl", "glu2hr_mgdl", "hba1c"), function(i) {
  do.call(rbind.data.frame, lapply(taste_diplos, function(d) {
    print_lm(outcome = i, exposure="taste_diplos", label=d[[1]], covariates = "age+sex",
             data=analysis %>% mutate(taste_diplos = relevel(taste_diplos, ref=d)))
  })) 
})%>% kable() 

```

```{r, fig.width=16, fig.height=8}
## Additional plot of glucose by fasting time -------
analysis %>% filter(complete.cases(taste_diplos)) %>% 
  ggplot(aes(x=as.factor(fasting_hrs), y=glu, color=taste_diplos, fill=taste_diplos)) +
  geom_boxplot(position=position_dodge(0.85), width=0.65) + 
  scale_color_manual(values=palettes$nature_diplos_dominant, name="TAS2R38 Diplotype") + 
  scale_fill_manual(values=palettes$nature_diplos_dominant, name="TAS2R38 Diplotype")


## Additional plot of glucose by fasting time -------
#analysis %>% filter(complete.cases(taste_diplos)) %>% 
  #mutate(fast_cat_v2 = case_when(fasting_hrs %in% c(6, 7, 8) ~ "6to8hrs"
 # ggplot(aes(x=as.factor(fasting_hrs), y=glu, color=taste_diplos, fill=taste_diplos)) +
#  geom_boxplot(position=position_dodge(0.85), width=0.65) + 
 # scale_color_manual(values=palettes$nature_diplos_dominant, name="TAS2R38 Diplotype") + 
#  scale_fill_manual(values=palettes$nature_diplos_dominant, name="TAS2R38 Diplotype")


#lm_gluxfasthr <- summary(lm(formula(paste0("glu~taste_diplos.num*as.factor(fasting_hrs)+", 
#                          gsub("fast_cat","as.factor(fasting_hrs)", models.l$BMI))), 
#           data=analysis))$coef %>% as.data.frame() %>%
#  mutate(Exposure=gsub("as.factor[(]","",gsub("[)]","", gsub("[.]num[:]", "*", rownames(.)))), .before=Estimate) %>%
#  filter(startsWith(Exposure, "taste")) %>%
#  rename(Beta=Estimate, SE=`Std. Error`, T.stat=`t value`, P=`Pr(>|t|)`) ; rownames(lm_gluxfasthr) <- NULL
#lm_gluxfasthr %>% kable()
```

***

### Associations of TAS2R38 diplotype with glycemic measures

We used generalized linear models with 4 models.
* Exposure: categorical diplotype (reference = AVI/AVI)
* Outcomes: continunous glycemic measures
* Covariates: 4 models; base, bmi, lifestyle, food choices


```{r, error=F, out.width="90%"}
AAs <- c(" AVI/AVI", "  AVI/PAV", "  PAV/PAV")

## Print table of associations of Diplos with RG ---------------------------
make_pretty_lm(read.csv(paste0("../data/processed/res_tab_lm_rg_tastediplo_mgdl_", tag,".csv")), 
               show_SE = T, digits=c(4,6)) %>% 
  kable(caption="Associations of TAS2R38 diplotype with Random Glucose")

make_pretty_lm(read.csv(paste0("../data/processed/res_tab_lm_rg_tastediplo_add_mgdl_", tag,".csv")), 
               show_SE = T, digits = c(4,6)) %>% 
  kable(caption="Associations of continuous TAS2R38 diplotype with Random Glucose")


## Print table of associations of Diplos with 2hrGlu ---------------------------
make_pretty_lm(read.csv(paste0("../data/processed/res_tab_lm_2hg_tastediplo_mgdl_", tag, ".csv")), 
               show_SE = T, digits = c(4,3)) %>%
  kable(caption="Associations of TAS2R38 diplotype with 0-2hr Glucose")

make_pretty_lm(read.csv(paste0("../data/processed/res_tab_lm_2hg_tastediplo_add_mgdl_", tag, ".csv")),  show_SE = T) %>%
  kable(caption="Associations of continuous TAS2R38 diplotype with 0-2hr Glucose")


## Print table of associations of Diplos with A1c ---------------------------
make_pretty_lm(read.csv(paste0("../data/processed/res_tab_lm_a1c_tastediplo_",tag,".csv")),  show_SE = T, digits = c(4,6)) %>%
  kable(caption="Associations of TAS2R38 diplotype with HbA1c")

```


##### Plot of estimated marginal means 
```{r}
## Glucose stratified by fasting time in all models ---------------------------
make_pretty_lm(read.csv(paste0("../data/processed/res_tab_lm_gluXfast12to24_tastediplo_mgdl_",tag,".csv"))) %>%
  kable(caption="Associations of TAS2R38 with Glucose, Stratified by Fasting Window")

# load emm data
emm_gluXfast <- read.csv(paste0("../data/processed/res_tab_emm_gluXfast_tastediplo_mDiet_mgdl_",tag,".csv"))
emm_gluXfast_6to12 <- read.csv(paste0("../data/processed/res_tab_emm_gluXfast_6to12_tastediplo_mDiet_nofast_mgdl_",tag,".csv"))
emm_allglycemic <- read.csv(paste0("../data/processed/res_tab_emm_allglycemic_tastediplos_mgdl_", tag, ".csv"))

# NOTE: looking deeper into the 12-24 range showed a TON of variability, so the EMMs just focused on the 6-12 window
emm_gluXfast %>%
  mutate(fast = factor(fast, levels=c("6+hr", "5hr", "4hr", "3hr", "0to2hr"))) %>%
  ggplot(aes(y=emmean, x=fast, ymin=lowCI, ymax=upCI, color=level)) + 
  facet_wrap(~model, ncol=1, scale="free", labeller = as_labeller(c("Diet.Patterns"="Glucose by Self-Reported Time Since Last Meal"))) +
  coord_flip() + geom_point(size=1.715, position=position_dodge(0.5)) + 
  geom_errorbar(width=0.35, position=position_dodge(0.5), linewidth=0.65)
emm_gluXfast_6to12 %>%
  mutate(fast = factor(fast, levels=c("12+hr", "6to12hr", "5hr", "4hr", "3hr", "0to2hr"))) %>%
  ggplot(aes(y=emmean, x=fast, ymin=lowCI, ymax=upCI, color=level)) + 
  facet_wrap(~model, ncol=1, scale="free", labeller = as_labeller(c("Diet.Patterns"="Glucose by Self-Reported Time Since Last Meal"))) +
  coord_flip() + geom_point(size=1.715, position=position_dodge(0.5)) + 
  geom_errorbar(width=0.35, position=position_dodge(0.5), linewidth=0.65)
```

### Create combined panel plot of glycemic traits by TAS2R38 diplotype

```{r}
#glucose by fasting time
p_emm_gluXfast_diet <- emm_gluXfast_6to12 %>%
  filter(fast != "12+hr") %>%
  mutate(fast=factor(fast, levels=c("6to12hr", "5hr", "4hr", "3hr", "0to2hr"),
                     labels =c("6-12 hrs", "5 hrs", "4 hrs", "3 hrs", "0-2 hrs")),
         Diplotypes = factor(level, levels=c("PAV/PAV", "AVI/PAV", "AVI/AVI"))) %>%
  ggplot(aes(y=emmean, x=fast, ymin=lowCI, ymax=upCI, color=Diplotypes)) + 
  facet_wrap(~model, ncol=1, scale="free", labeller = as_labeller(c("Diet.Patterns"="Glucose by Self-Reported Fasting Time"))) +
  coord_flip() + scale_y_continuous(limits=c(4.81, 4.965), breaks=seq(4.80, 4.96,0.02)) +
  geom_point(size=1.715, position=position_dodge(0.5)) + 
  geom_errorbar(width=0.35, position=position_dodge(0.5), linewidth=0.65) + 
  scale_color_manual(values=palettes$nature_diplos_dominant, name="TAS2R38 Diplotype") +
  ylab("Blood glucose, mmol/L") + xlab(" ") + 
  ggtheme_figures + ggtheme_emm + theme(
    strip.text = element_text(face="bold", size=11),
    strip.background = element_rect(color="black")) ; p_emm_gluXfast_diet

## random glucose 
p_emm_glu <- emm_allglycemic %>%
  filter(outcome == "Glucose (mg/dL)" & model == "Diet.Patterns") %>%
  mutate(Diplotypes = factor(level, levels=c("PAV/PAV", "AVI/PAV", "AVI/AVI")),
         glycemic = factor(outcome, levels=c("Glucose", "HbA1c"))) %>%
  ggplot(aes(y=emmean, x=Diplotypes, ymin=lowCI, ymax=upCI, color=Diplotypes)) + 
  facet_wrap(~glycemic, ncol=1, scale="free",labeller=as_labeller(c("Glucose"="Random Glucose"))) +
  coord_flip() + scale_y_continuous(limits=c(4.855, 4.885)*18.018, breaks=seq(4.855, 4.885,0.01)*18.018) +
  geom_point(size=1.715, position=position_dodge(0.75)) + 
  geom_errorbar(width=0.35, position=position_dodge(0.75), linewidth=0.65) + 
  scale_color_manual(values=palettes$nature_diplos_dominant, name="TAS2R38 Diplotype") +
  ylab("Blood glucose, mmol/L") + xlab(" ") + 
  ggtheme_figures + ggtheme_emm + theme(
    strip.text = element_text(face="bold", size=11), 
    strip.background = element_rect(color="black")) ; p_emm_glu

#a1c
p_emm_a1c <- emm_allglycemic %>%
  filter(outcome == "HbA1c (%)" & model == "Diet.Patterns") %>%
  mutate(Diplotypes = factor(level, levels=c("PAV/PAV", "AVI/PAV", "AVI/AVI")),
         glycemic = factor(outcome, levels=c("Glucose", "HbA1c"))) %>%
  ggplot(aes(y=emmean, x=Diplotypes, ymin=lowCI, ymax=upCI, color=Diplotypes)) + 
  facet_wrap(~glycemic, ncol=1, scale="free") +
  coord_flip() + scale_y_continuous(limits=c(5.281, 5.291), breaks=seq(5.28, 5.291,0.002)) +
  geom_point(size=1.715, position=position_dodge(0.75)) + 
  geom_errorbar(width=0.35, position=position_dodge(0.75), linewidth=0.65) + 
  scale_color_manual(values=palettes$nature_diplos_dominant, name="TAS2R38 Diplotype") +
  ylab("HbA1c, mmol/L") + xlab(" ") + 
  ggtheme_figures + ggtheme_emm + theme(
    strip.text = element_text(face="bold", size=11),
    strip.background = element_rect(color="black")) ; p_emm_a1c

## PANEL plot    
emm_panel_allglycemic <- ggarrange(
  ggarrange(p_emm_glu, "", p_emm_a1c, heights=c(0.75,0.15,0.75), labels=c("A", "", "B"),
            common.legend = T, legend = "none", ncol=1),
  p_emm_gluXfast_diet, ncol=2, common.legend = T, legend="bottom", widths=c(0.8, 1.1), 
  labels=c("", "C")) ; emm_panel_allglycemic

#emm_panel_allglycemic %>% ggsave(filename=paste0("../output/plot_emm_allglycemic_tastediplos_mDiet_",tag,".pdf"), height=4.75, width=9.6)
```

## V2 - combining random & fasting-stratified glucose levels
```{r}
#glucose by fasting time
emm_glu_mgdl_diet <- rbind.data.frame(
  emm_allglycemic %>% filter(outcome == "Glucose (mg/dL)" & model == "Diet.Patterns") %>% 
    mutate(fast = "Random", .after="model"), emm_gluXfast_6to12)

p_emm_gluXfast_diet <- emm_glu_mgdl_diet %>%
  filter(fast != "12+hr") %>%
  mutate(fast=factor(fast, levels=rev(c("6to12hr", "5hr", "4hr", "3hr", "0to2hr", "Random")),
                     labels=rev(c("6-12 hrs", "5 hrs", "4 hrs", "3 hrs", "0-2 hrs", "Anytime from\n0-12 hrs"))),
         Diplotypes = factor(level, levels=c("AVI/AVI", "AVI/PAV", "PAV/PAV"))) %>%
  mutate(type = factor(ifelse(fast=="Anytime from\n0-12 hrs", "Random", "Fasting"), levels=c("Random", "Fasting"))) %>%
  ggplot(aes(y=emmean, x=fast, ymin=lowCI, ymax=upCI, color=Diplotypes)) + 
  facet_grid(~type, scale="free", space="free_x", labeller = as_labeller(c("Random" = "Random Glucose", "Fasting"="Glucose by Self-Reported Fasting Time"))) +
  scale_y_continuous(limits=c(86.75, 89.7), breaks=seq(87, 89.7,0.5)) +
  geom_point(size=1.715, position=position_dodge(0.5)) + 
  geom_errorbar(width=0.35, position=position_dodge(0.5), linewidth=0.65) + 
  scale_color_manual(values=palettes$nature_diplos_dominant, name="TAS2R38 Diplotype") +
  ylab("Glucose level (mg/dL)\n") + xlab(" ") + 
  ggtheme_basic +
 theme(
    legend.position = "bottom", legend.text = element_text(size=10),
    legend.title = element_text(size=10), 
   # axis.ticks.x=element_blank(),
    panel.grid.major.y = element_line(), 
    #strip.background = element_blank(), 
    panel.background = element_rect(fill="#00000010")
    ) ; p_emm_gluXfast_diet

#p_emm_gluXfast_diet %>% ggsave(filename=paste0("../output/plot_emm_allglu_tastediplos_mDiet_mgdl_",tag,".pdf"), height=4.75, width=7.15)
p_emm_gluXfast_diet %>% ggsave(filename=paste0("../output/plot_emm_allglu_tastediplos_mDiet_mgdl_",tag,"_BR.pdf"), height=3.75, width=7)

```

```{r}

plot_emm_pvals <- c("P=0.051", "P=0.016", "P=0.798", "P=0.454", "P=0.692", "P=0.056")

p_emm_gluXfast_diet_wh <- emm_glu_mgdl_diet %>%
  filter(fast != "12+hr") %>%
  mutate(fast=factor(fast, levels=rev(c("6to12hr", "5hr", "4hr", "3hr", "0to2hr", "Random")),
                     labels=rev(c("6-12 hr", "5 hr", "4 hr", "3 hr", "0-2 hr", "Anytime from\n0-12 hrs"))),
         Diplotypes = factor(level, levels=c("AVI/AVI", "AVI/PAV", "PAV/PAV"))) %>%
  mutate(type = factor(ifelse(fast=="Anytime from\n0-12 hrs", "Random", "Fasting"), levels=c("Random", "Fasting"))) %>%
  ggplot(aes(y=emmean, x=fast, ymin=lowCI, ymax=upCI, color=Diplotypes)) + 
  #facet_grid(~type, scale="free", space="free_x", labeller = as_labeller(c("Random" = "Random Glucose", "Fasting"="Glucose by Self-Reported Fasting Time"))) +
  facet_grid(~type, scale="free", space="free_x", labeller = as_labeller(c("Random" = "", "Fasting"=""))) +
  scale_y_continuous(limits=c(86.7, 90), breaks=seq(87, 90,0.5)) +
  geom_point(size=2.25, position=position_dodge(0.55)) + 
  geom_errorbar(width=0.35, position=position_dodge(0.55), linewidth=0.85) + 
  scale_color_manual(values=palettes$nature_diplos_dominant, name="**_TAS2R38_ Diplotype**") +
  ylab("Glucose level (mg/dL)\n") + xlab("Self-reported fasting window (hr)") + 
  theme_bw() +
  theme(#legend.position = "right",
        axis.title.x = element_text(size=12, face="bold", vjust=2), 
        axis.title.y = element_text(size=12, face="bold", vjust=-2), 
        axis.text.x = element_text(size=10), axis.text.y = element_text(size=10),
        panel.grid.major.y = element_line(), 
        strip.background = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.background = element_rect(fill="#00000000"),
        strip.text = element_text(face="bold", size=10), 
        
        legend.position = "inside",
        legend.title = element_markdown(), #element_text(face="bold", size=9),
        legend.text = element_text(size=10, face="italic"), #legend.title=element_blank(),
        legend.position.inside = c(0.99, 0.98),  # Moves legend to top-right inside plot
        legend.justification.inside = c(1, 1),  # Aligns legend's top-right corner
        legend.background = element_rect(fill = "white", color = "black", linewidth=0.15), 
          
        legend.key.size = unit(1, 'line'),
        legend.margin = margin(3,3,3,3, unit="pt") 
  )

p_emm_gluXfast_diet_wh
p_emm_gluXfast_diet_wh %>% ggsave(filename=paste0("../output/plot_emm_allglu_tastediplos_mDiet_mgdl_",tag,"_wh_vManuscript.pdf"), height=3.75, width=8)


```

#### Additional analysis: TAS2R38 with T2D status
```{r}
fread("../data/processed/res_tab_glm_t2d_tastediplo_v6.csv")

```
#### End of primary analysis tables & figures 





